---
title: "Examining correlation structure in RNA-seq data"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/combined_enrichment

## Intro

This script is designed to fetch Illumina Human Body Map 2.0 Project gene
expression data in order to assess the correlation of genes within a gene set.

The data is being fetched using getDEE2 with the accession number ERP000546.
```{r,begin0}

suppressPackageStartupMessages({
  library("kableExtra")
  library("beeswarm")
  library("getDEE2")
  library("mitch")
  library("gplots")
  library("Hmisc")
  library("parallel")
  library("vioplot")
})

```

## Fetch data

First get gene expression data.

```{r,fetch_expression_data}

mdat <- getDEE2Metadata(species="hsapiens")
mdat2 <- mdat[which(mdat$SRP_accession == "ERP000546"),]
dat <- getDEE2::getDEE2(species="hsapiens",
  metadata=mdat,
  SRRvec=mdat2$SRR_accession,
  legacy=TRUE)

dat <- Tx2Gene(dat)
names(dat)
head(dat$Tx2Gene)
head(dat$GeneInfo)
tx2gene <- dat$Tx2Gene
rownames(tx2gene) <- sapply(strsplit(rownames(tx2gene),"\\."),"[[",1)

# filter based on detection threshold
tx2gene <- tx2gene[which(rowMeans(tx2gene)>=10),]

# normalise
rpm <- tx2gene/colSums(tx2gene) * 1000000

```

Next get pathways.

```{r,reactome}

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
genesets<-gmt_import("ReactomePathways.gmt")

```

## Assess the correlation structure

```{r,flatcor}

# ++++++++++++++++++++++++++++
# flattenCorrMatrix
# ++++++++++++++++++++++++++++
# cormat : matrix of the correlation coefficients
# pmat : matrix of the correlation p-values
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

```


```{r,cor1}


res <- mclapply(1:length(genesets), function(i){
gs <- genesets[[i]]
len <- length(gs)
gsname <- names(genesets[i])
gene_accessions <- rownames(dat$GeneInfo[which(dat$GeneInfo$GeneSymbol %in% gs),])
x <- rpm[which(rownames(rpm) %in% gene_accessions),]
if(nrow(x)>=10) {
  res <- rcorr(as.matrix(t(x)),type="spearman")
  res <- flattenCorrMatrix(res$r, res$P)
  # randomly select gene ids as a comparison
  gene_accessions2 <- rownames(dat$GeneInfo[which(! dat$GeneInfo$GeneSymbol %in% gs),])
  # filter for those expressed in the dataset and randomly select n
  gene_accessions2 <- gene_accessions2[sample(which(gene_accessions2 %in% rownames(rpm) ),size=len, replace=FALSE)]
  y <- rpm[which(rownames(rpm) %in% gene_accessions2),]
  res2 <- rcorr(as.matrix(t(y)),type="spearman")
  res2 <- flattenCorrMatrix(res2$r, res2$P)
  cormax <- max(c(res$cor,res2$cor))
  cormin <- min(c(res$cor,res2$cor))
  setname <- names(genesets[i])
  result <- list("inset"=res$cor, "rand"=res2$cor,"setname"=setname)
  return(result) 
}
}, mc.cores=16)

```

looks like some empty results. Let's get rid of those now.

```{r,filt}

str(head(res))
res <- res[which(lapply(res,length)>0)]
str(head(res))

```

We can visualise the spearman correlation coefficients here.

```{r,viz1}

z <- lapply(1:6, function(i){
setname <- res[[i]]$setname
boxplot(res[[i]][1:2],col="white",ylab="Spearman rho",main=setname)
beeswarm(res[[i]][1:2],add=TRUE,pch=19)
})

```

We can run a t-test to quantify the difference in distributions.

```{r,test1}

res2 <- lapply(1:length(res), function(i) {
inset <- res[[i]]$inset
rand <- res[[i]]$rand
tt <- t.test(x=inset,y=rand)
return(c("stat"=tt$statistic,"pvalue"=tt$p.value))
})

res2 <- do.call(rbind,res2)

rownames(res2) <- sapply(res,"[[",3)
summary(res2)

```

Now we can look at the positive and negatively correlated gene sets.

```{r,poscor}

poscor <- head(res2[order(-res2[,1]),],10) 
poscor %>% kbl(caption="most correlated gene sets") %>% kable_paper("hover", full_width = F)
poscor_res <- res[which(sapply(res,"[[",3) %in% rownames(poscor))]

```

Vioplot of positively correlated sets.

```{r,posvio}

z <- lapply(1:length(poscor_res), function(i){
setname <- poscor_res[[i]]$setname
gset <- genesets[[which(names(genesets) ==setname)]]
eset <- rownames(dat$GeneInfo[which(dat$GeneInfo$GeneSymbol %in% gset),])
n <- length(which(eset %in% rownames(rpm)))
vioplot(poscor_res[[i]][1:2] , ylab="Spearman rho",main=setname  )
mtext(paste("n=",n))
})

```

Heatmap of pos cor sets.

```{r,posheat}
z <- lapply(1:10, function(i) {
setname <- rownames(poscor)[i]
gset <- genesets[[which(names(genesets) ==setname)]]
eset <- rownames(dat$GeneInfo[which(dat$GeneInfo$GeneSymbol %in% gset),])
mx <- rpm[which(rownames(rpm) %in% eset),]
heatmap.2(as.matrix(mx),trace="none",scale="row",main=setname, margin=c(5,7))
heatmap.2(cor(t(mx)),trace="none",scale="none",main=setname,margin=c(5,7))
})

```

Now focus on negatively correlated sets.

```{r,anticor}

anticor <- head(res2[order(res2[,1]),],10)
anticor %>% kbl(caption="most anti-correlated gene sets") %>% kable_paper("hover", full_width = F)
anticor_res <- res[which(sapply(res,"[[",3) %in% rownames(anticor))]

```

```{r,negvio}

z <- lapply(1:length(anticor_res), function(i){
setname <- anticor_res[[i]]$setname
gset <- genesets[[which(names(genesets) ==setname)]]
eset <- rownames(dat$GeneInfo[which(dat$GeneInfo$GeneSymbol %in% gset),])
n <- length(which(eset %in% rownames(rpm)))
vioplot(anticor_res[[i]][1:2] , ylab="Spearman rho",main=setname  )
mtext(paste("n=",n))
})

```

Heatmap of neg cor sets.

```{r,negheat}

z <- lapply(1:10, function(i) {
setname <- rownames(anticor)[i]
gset <- genesets[[which(names(genesets) ==setname)]]
eset <- rownames(dat$GeneInfo[which(dat$GeneInfo$GeneSymbol %in% gset),])
mx <- rpm[which(rownames(rpm) %in% eset),]
heatmap.2(as.matrix(mx),trace="none",scale="row",main=setname,margin=c(5,7))
heatmap.2(cor(t(mx)),trace="none",scale="none",main=setname,margin=c(5,7))
})

```
Now visualise the test statistics found.

```{r,viz2}

par(mfrow=c(2,1))
hist(res2[,"stat.t"],breaks=100,main="",xlab="test statistic")
boxplot(res2[,"stat.t"],horizontal=TRUE)


hist(res2[,"stat.t"],xlim=c(-50,50),breaks=100,main="",xlab="test statistic")
boxplot(res2[,"stat.t"],ylim=c(-50,50),horizontal=TRUE)

par(mfrow=c(1,1))

pos <- length( which(res2[,1]>0 & res2[,2] < 0.05) )
neg <- length( which(res2[,1]<0 & res2[,2] < 0.05) )
none <- length( which(res2[,2] > 0.05) )
bars <- c("Pos"=pos,"None"=none,"Neg"=neg)
barplot(bars,ylab="no. gene sets with correlated expression",ylim=c(0,max(bars)*1.3))
text(((1:length(bars))-0.5)*1.25,bars+50,labels=bars)

```

## Conclusions

This work confirms the observations by Hong, that genes in a set are on the whole, most gene sets positively correlated members.

Some gene sets are highly correlated, while others are heterogenous, or slightly anticorrelated.

We can try to reproduce this correlation structure in simulated data.

## Session information

```{r,session0}

sessionInfo()

```
