---
title: "Example gene set analysis: The case of HDAC inhibitor SAHA"
author: "Mark Ziemann & Kaumadi Wijesooriya"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
theme: cosmo
---

Source: https://github.com/markziemann/combined_enrichment

## Intro

Here we are performing an analysis of some gene expression data to demonstrate the difference between ORA and FCS methods and to highlight the differences caused by improper background gene set use.

The dataset being used is SRP096178 and we are comparing the cells grown in normal condition (control) to those grown with addition of SAHA (case).

Data are obtained from http://dee2.io/

```{r,begin}

suppressPackageStartupMessages({
library("getDEE2") 
library("DESeq2")
library("clusterProfiler")
library("mitch")
library("kableExtra")
library("eulerr")
})

```

## Get expression data

I'm using some RNA-seq data looking at the effect of SAHA on HAEC cells.

```{r,getdata,fig.width=7,fig.height=7}

name="SRP037718"
mdat<-getDEE2Metadata("hsapiens")
samplesheet <- mdat[grep("SRP037718",mdat$SRP_accession),]
samplesheet<-samplesheet[order(samplesheet$SRR_accession),]

samplesheet$trt<-as.factor(c(1,1,1,0,0,0))
s1 <- samplesheet

s1 %>% kbl(caption = "sample sheet") %>% kable_paper("hover", full_width = F)

w<-getDEE2("hsapiens",s1$SRR_accession,metadata=mdat,legacy = TRUE)
x<-Tx2Gene(w)
x<-x$Tx2Gene

# save the genetable for later
gt<-w$GeneInfo[,1,drop=FALSE]
gt$accession<-rownames(gt)

# counts 
x1<-x[,which(colnames(x) %in% samplesheet$SRR_accession)]

```

Here show the number of genes in the annotation set, and those detected above the detection threshold.

```{r,filter}

# filter out lowly expressed genes
x1<-x1[which(rowSums(x1)/ncol(x1)>=(10)),]
nrow(x)
nrow(x1)

```

Now multidimensional scaling (MDS) plot to show the correlation between the datasets.
If the control and case datasets are clustered separately, then it is likely that there will be many differentially expressed genes with FDR<0.05.

```{r,mds}

plot(cmdscale(dist(t(x1))), xlab="Coordinate 1", ylab="Coordinate 2", pch=19, col=s1$trt, main="MDS")

```

## Differential expression

Now run DESeq2 for control vs case.

```{r,deseq2}

y <- DESeqDataSetFromMatrix(countData = round(x1), colData = s1, design = ~ trt)
y <- DESeq(y)
de <- results(y)
de<-as.data.frame(de[order(de$pvalue),])
rownames(de)<-sapply(strsplit(rownames(de),"\\."),"[[",1)
head(de) %>% kbl() %>% kable_paper("hover", full_width = F)

```

Now let's have a look at some of the charts showing differential expression.
In particular, an MA plot and volcano plot.

```{r,deplots,fig.width=7,fig.height=7}

maplot <- function(de,contrast_name) {
  sig <-subset(de, padj < 0.05 )
  up <-rownames(subset(de, padj < 0.05 & log2FoldChange > 0))
  dn <-rownames(subset(de, padj < 0.05 & log2FoldChange < 0))
  GENESUP <- length(up)
  GENESDN <- length(dn)
  DET=nrow(de)
  SUBHEADER = paste(GENESUP, "up, ", GENESDN, "down", DET, "detected")
  ns <-subset(de, padj > 0.05 )
  plot(log2(de$baseMean),de$log2FoldChange, 
       xlab="log2 basemean", ylab="log2 foldchange",
       pch=19, cex=0.5, col="dark gray",
       main=contrast_name, cex.main=0.7)
  points(log2(sig$baseMean),sig$log2FoldChange,
         pch=19, cex=0.5, col="red")
  mtext(SUBHEADER,cex = 0.7)
}

make_volcano <- function(de,name) {
    sig <- subset(de,padj<0.05)
    N_SIG=nrow(sig)
    N_UP=nrow(subset(sig,log2FoldChange>0))
    N_DN=nrow(subset(sig,log2FoldChange<0))
    DET=nrow(de)
    HEADER=paste(N_SIG,"@5%FDR,", N_UP, "up", N_DN, "dn", DET, "detected")
    plot(de$log2FoldChange,-log10(de$padj),cex=0.5,pch=19,col="darkgray",
        main=name, xlab="log2 FC", ylab="-log10 pval", xlim=c(-6,6))
    mtext(HEADER)
    grid()
    points(sig$log2FoldChange,-log10(sig$padj),cex=0.5,pch=19,col="red")
}

maplot(de,name)
make_volcano(de,name)

```

## Gene sets from Reactome

In order to perform gene set analysis, we need some gene sets. 

```{r,reactome}

if (! file.exists("ReactomePathways.gmt")) {
  download.file("https://reactome.org/download/current/ReactomePathways.gmt.zip", 
    destfile="ReactomePathways.gmt.zip")
  unzip("ReactomePathways.gmt.zip")
}
genesets<-gmt_import("ReactomePathways.gmt")

```


## FCS with Mitch

Mitch uses rank-ANOVA statistics for enrichment detection.

Here I'm using the standard approach

```{r,mitch1}

m <- mitch_import(de,DEtype = "DEseq2", geneTable = gt)
msep <- mitch_calc(m,genesets = genesets)
ms_up <- subset(msep$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
ms_dn <- subset(msep$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
message(paste("Number of up-regulated pathways:",length(ms_up) ))
message(paste("Number of down-regulated pathways:",length(ms_dn) ))
head(msep$enrichment_result,10)  #%>% kbl() %>% kable_paper("hover", full_width = F)
#mitch_report(msep,outfile="mitch_separate.html",overwrite=TRUE)

```

Here I'm using the combined approach.

```{r,mitch2}

mcom <- mitch_calc(abs(m),genesets = genesets)
mc_up <- subset(mcom$enrichment_result,p.adjustANOVA<0.05 & s.dist > 0)[,1]
mc_dn <- subset(mcom$enrichment_result,p.adjustANOVA<0.05 & s.dist < 0)[,1]
message(paste("Number of up-regulated pathways:",length(mc_up) ))
message(paste("Number of down-regulated pathways:",length(mc_dn) ))
head(mcom$enrichment_result,10)  #%>% kbl() %>% kable_paper("hover", full_width = F)
#mitch_report(mcom,outfile="mitch_combined.html",overwrite=TRUE)

```

Let's look at the significant ones based on the combined analysis.
There weren't many gene sets classe
d as significant.
Let's see how many have a direction agnostic enrichment score which is larger in magnitude than the direction informed enrichment score.
There are only 11 such sets which would benefit from such combined analysis.

Euler diagram of the significant pathways found with each approach.

```{r,mitch_euler1}

l0 <- list("sep up"=ms_up,"sep dn"=ms_dn,"comb up"=mc_up,"comb dn"=mc_dn)
par(cex.main=0.5)
plot(euler(l0),quantities = TRUE, edges = "gray", main="FCS: combined vs separated")


length(ms_up)
length(ms_dn)
length(ms_up)+length(ms_dn)

length(mc_up)
length(mc_dn)
length(mc_up)+length(mc_dn)

( length(ms_up)+length(ms_dn) ) / ( length(mc_up)+length(mc_dn) )


```

List gene sets which are specific to each approach.

```{r,mitch_sets}

ms <- c(ms_up,ms_dn)

# in sep but not comb
setdiff(ms,mc_up)

# in comb but not sep
setdiff(mc_up,ms)

# intersection
intersect(mc_up,ms)

```

If we consider both strategies to be valid, then we can define the significant sets as
dysregulated. 
We can calculate the percent sentitivity of both approaches.

```{r,mitch_proportion}

all <- unique(c(ms_up,ms_dn,mc_up))

message("Sensitivity: separate only")
(length(ms_up)+length(ms_dn))/length(all)

message("Sensitivity: combined only")
length(mc_up)/length(all)

```

## ORA with clusterprofiler

Clusterprofiler uses a hypergeometric test.
Firstly I will conduct the analysis separately for up and down regulated genes and with the correct backgound (as intended by the developers).

```{r,cp1}

genesets2 <- read.gmt("ReactomePathways.gmt")

de_up <- rownames(subset(de, padj<0.05 & log2FoldChange > 0))
de_up <- unique(gt[which(rownames(gt) %in% de_up),1])

de_dn <- rownames(subset(de, padj<0.05 & log2FoldChange < 0))
de_dn <- unique(gt[which(rownames(gt) %in% de_dn),1])

de_bg <- rownames(de)
de_bg <- unique(gt[which(rownames(gt) %in% de_bg),1])

o_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
o_up <- rownames(subset(o_up, p.adjust < 0.05))
       
o_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
o_dn <- rownames(subset(o_dn, p.adjust < 0.05))

o_com <- as.data.frame(enricher(gene = union(de_up,de_dn), universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
o_com <- rownames(subset(o_com, p.adjust < 0.05))

length(o_up)
length(o_dn)
length(o_up) + length(o_dn)
length(o_com)

( length(o_up) + length(o_dn) ) / length(o_com)

all <- unique(c(o_up,o_dn,o_com))

message("Sensitivity: separate only")
(length(o_up)+length(o_dn))/length(all)

message("Sensitivity: combined only")
length(o_com)/length(all)

```

Euler diagram of the significant pathways found with each approach.

```{r,cp_euler1}

l2 <- list("sep up"=o_up,"sep dn"=o_dn,"comb"=o_com)

plot(euler(l2),quantities = TRUE, edges = "gray", main="ORA: combined vs separated")

```

List gene sets which are specific to each approach.

```{r,ora_sets}

o_sep <- c(o_up,o_dn)

# in sep but not comb
setdiff(o_sep,o_com)

# in comb but not sep
setdiff(o_com,o_sep)

# intersection
intersect(mc_up,ms)

```

## Euler diagrams comparing FCS and ORA methods

```{r,euler_comparison}

par(cex.main=0.5)

par(mar=c(2,2,2,2))

l3 <- list("ORA up"=o_up,"ORA dn"=o_dn,"ORA comb"=o_com,
  "FCS up"=ms_up,"FCS dn"=ms_dn,"FCS comb"=mc_up)

plot(euler(l3),quantities = TRUE, edges = "gray", main="FCS compared to ORA")

```

## Save data

```{r,save}

dat <- list(  "FCS_up"=ms_up,
  "FCS_dn"=ms_dn,
  "FCS_com"=mc_up,
  "ORA_up"= o_up,
  "ORA_dn"=o_dn,
  "ORA_com"=o_com)

str(dat)
saveRDS(dat,file = "ex3dat.rds")

```

## Conclusion

For mitch, it would appear that performing direction informed (DI) analysis clearly yields more differentially regulated pathways
(413) as compared to the direction agnostic (DA) method (55).

That being said, there were 18 pathways identified only in the DA method that appeared to be related to the physiology of the model.
These gene sets are likely to contain a mix of genes affected by the stimulus in different ways - for example a mix of up and 
downregulated genes.
Are these really real? Not sure.

This pattern was consistent with ORA, where 80 sets were identified with separate analysis and only 23 with the combined analysis.

When comparing ORA to FCS, we found that FCS identified many more sets than ORA. 
In fact all gene sets that were identified by ORA were also identified by FCS, except for 3 that were specific to the ORA up set.

Let's look at those now.

```{r,conc}

myfcs <- c(ms_up, mc_up)

setdiff(o_up,myfcs)

```

## Session information

```{r,session}

sessionInfo()

```
